Chess EDA

### 2023-03-28 owen/01_eda.qmd
some more stuff for james

@@

mcmc_hist(fit3_ave$draws(c("beta", "gamma1", "gamma2")))
```

@@

@@

@@

```{r}
#| eval: false

stan_data <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$focal_win_prop)

stan_file <- "model1.stan"

fit1 <- mod$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit1$summary()

## but seems to give positive winner effects for all players
mcmc_hist(fit1$draws("alpha"))
mcmc_hist(fit1$draws("beta"))

# fit$save_object(file = "simple_multi_logistic.RDS")

@@

```{r}
#| eval: false
stan_file <- "model2.stan"

mod2 <- cmdstan_model(stan_file)

fit2 <- mod2$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit2$summary()

mcmc_hist(fit2$draws(c("beta", "gamma1", "gamma2")))

# fit$save_object(file = "cov_multi_logistic.RDS")

## these indicate clear positive coefficients for the betas,
## which are basically winner effects, account for the role
## of win history
## elo seems to have a much smaller role

## repeat for a different history, did this using past 5 games, could try
## others.
## could also compare to longer games, see if you get the same thing
## don't put a common prior on the betas, which is what the Cribben paper
## did. also don't standardize them

```{r}
#| eval: false
stan_file <- "model3.stan"

mod3 <- cmdstan_model(stan_file)

fit3 <- mod3$sample(data = stan_data,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit3$summary()
mcmc_hist(fit3$draws(c("beta", "mu", "tau", "gamma1", "gamma2")))

mcmc_trace(fit3$draws(c("mu", "tau", "beta")))

```{r}
#| eval: false
stan_data_ave <- list(N = nrow(init_data),
                  J = length(top_players),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

stan_file <- "model1.stan"

fit1_ave <- mod$sample(data = stan_data_ave,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit1_ave$summary()

## but seems to give positive winner effects for all players
mcmc_hist(fit1_ave$draws("alpha"))
mcmc_hist(fit1_ave$draws("beta"))

```{r}
#| eval: false
stan_file <- "model2.stan"

fit2_ave <- mod$sample(data = stan_data_ave,
                  seed = 123,
                  chains = 4,
                  parallel_chains = 4,
                  refresh = 100)

fit2_ave$summary()

mcmc_hist(fit2$draws(c("beta", "gamma1", "gamma2")))
```

```{r}
#| eval: false
stan_file <- "model3.stan"

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "tau1", "gamma1", "gamma2")))
mcmc_hist(fit3_ave$draws(c("alpha", "mu2", "tau2")))
```

<!-- ## Streaks -->

>>>>

We first consider instead using the previous 5 games.

@@

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "gamma1", "gamma2")))
```

We also consider the previous 20 games.

@@

Finally, we consider the previous 50 games also.

@@

### Longer Games

In the above example we considered only rated 60 second bullet games.
We will now instead consider rated 3 minute blitz games, and fit
the same model. This would indicate how the effect might change as
we vary the length of the games considered. We have 12k total such blitz games.
We will use the same players as were considered for the bullet data, 
of which 16 play blitz games and only 7 play more than 100 blitz games.
This leads to 9000 games.

```{r get blitz data}

blitz_180 <- lichess_data %>% 
  filter(Event == "Rated Blitz game") %>% 
  filter(TimeControl == "180+0")

# blitz_180

blitz_top <- blitz_180 %>% 
  filter(Username %in% top_players) %>% 
  group_by(Username) %>% 
  tally() %>% 
  arrange(-n) %>% 
  pull(Username)

## will slightly change the order then compared to the previous data

tidy_games <- map_dfr(blitz_top, get_hist, blitz_180, prev_n = 5) %>% 
  as_tibble()

We first consider using the previous 5 games as the history. We remove
draws from the outcomes, but use them for the history.

```{r get stan data}
#| eval: false

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, blitz_top)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = focal_win_prop - mean(focal_win_prop)) %>% 
  filter(focal_result != 0.5)

stan_data_ave <- list(N = nrow(init_data),
                  J = length(blitz_top),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

## save this fit

fit3_ave$save_object("init_model_blitz_n5.RDS")

```{r}
fit3_ave <- readRDS("init_model_blitz_n5.RDS")

We can also repeat this with the previous 20 blitz games as an example also.
We only consider 13 players now who have more than 20 blitz games from the 
original selection.

```{r prev 20 blitz}
#| eval: false

tidy_games <- map_dfr(blitz_top[1:13], get_hist, blitz_180, prev_n = 20) %>% 
  as_tibble()

init_data <- tidy_games %>% 
  mutate(WhiteElo = as.numeric(WhiteElo), 
         BlackElo = as.numeric(BlackElo)) %>% 
  mutate(focal_user = ifelse(focal_white == 1, White, Black)) %>% 
  mutate(elo_diff = ifelse(focal_white == 1, 
                           WhiteElo - BlackElo, BlackElo - WhiteElo)) %>% 
  mutate(focal_id = match(focal_user, blitz_top)) %>% 
  select(focal_user, focal_id, focal_white, 
         focal_win_prop, elo_diff, focal_result) %>% 
  group_by(focal_id) %>% 
  mutate(ave_prop = focal_win_prop - mean(focal_win_prop)) %>% 
  filter(focal_result != 0.5)

stan_data_ave <- list(N = nrow(init_data),
                  J = length(blitz_top),
                  y = init_data$focal_result,
                  id = init_data$focal_id,
                  colour = init_data$focal_white,
                  elo = init_data$elo_diff,
                  win_prop = init_data$ave_prop)

## save this fit

fit3_ave$save_object("init_model_blitz_n20.RDS")

```{r}
fit3_ave <- readRDS("init_model_blitz_n20.RDS")

mcmc_hist(fit3_ave$draws(c("beta", "mu1", "gamma1", "gamma2")))
```

When we consider more historic games the winner effects do appear to shrink,
as we saw in the bullet data. However, there seems to still be
very strong evidence for some effect. This model does not fit as 
well, but this can be fixed.

@@

## Streaks

One of the other interesting components of this data is the performance
of a player after they draw the previous game, compared to when they
won or lost the previous game. There are a small number of draws is
this data set that we can use to investigate this.
We just get the proportion of results in the next game, for both the 
bullet and blitz data.

```{r}
#| echo: true
prev_data <- bullet_60 %>% 
  group_by(Username) %>% 
  arrange(UTCDate, UTCTime) %>% 
  ## get focal result here also
  mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
  mutate(focal_result = case_when(
    (focal_white == 1 & Result == "1-0") ~ 1,
    (focal_white == 0 & Result == "0-1") ~ 1,
    (Result == "1/2-1/2") ~ 0.5,
    .default = 0
    )) %>% 
  mutate(prev_result = lag(focal_result)) %>% 
  select(White, Black, WhiteElo, BlackElo, focal_white:prev_result)

prev_data %>% 
  drop_na() %>%
  group_by(prev_result, focal_result) %>% 
  tally() %>% 
  group_by(prev_result) %>% 
  mutate(result_prop = (n/sum(n)))

## get some confidence intervals for these differences, might be enough
## evidence for a winner effect after winning, maybe not for a loser effect 
## after losing

## can also repeat this for the blitz data

prev_blitz <- blitz_180 %>% group_by(Username) %>% 
  arrange(UTCDate, UTCTime) %>% 
  ## get focal result here also
  mutate(focal_white = ifelse(Username == White, 1, 0)) %>% 
  mutate(focal_result = case_when(
    (focal_white == 1 & Result == "1-0") ~ 1,
    (focal_white == 0 & Result == "0-1") ~ 1,
    (Result == "1/2-1/2") ~ 0.5,
    .default = 0
    )) %>% 
  mutate(prev_result = lag(focal_result)) %>% 
  select(White, Black, WhiteElo, BlackElo, focal_white:prev_result)

prev_blitz %>% 
  drop_na() %>%
  group_by(prev_result, focal_result) %>% 
  tally() %>% 
  group_by(prev_result) %>% 
  mutate(result_prop = (n/sum(n)))

## see something similar with the blitz data, maybe not as strong but
## seems to be there

@@
